Stage-Splitting a Modal Language

### 2014-06-17 paper/semantics.tex
Added more and better discussion of the fexp example.

Fast exponentiation is a method for calculating $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]
It is often observed that the decomposition of the power depends only on the power at prior recursive levels.
For this reason, we can stage an implementation of fast exponentiation so that the power is available at the first stage and the base is available at the second stage.
This implies that the result is then available only at the second stage.

We implemented staged fast exponentiation in \lang~below:

\begin{lstlisting} 
let fexp (b : $int, e : int) : $int =
	if e == 0 then
		next{1}
	else if (e mod 2) == 0 then
		next{let x = prev{fexp(b,e/2)} in x*x}
	else
		next{prev{b} * prev{fexp(b,e-1)}}		
\end{lstlisting}

The type of the recursive function is $\fut{\rm int} \times {\rm int} \to \fut{\rm int}$.
Note that the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the first stage.

% FIXME but do we give a cost semantics? no
Given this definition of fast exponent, the question the remains how it can be run.  For this, we need a semantics for \lang.

>>>>

Fast exponentiation is a method for calculating $b^p$ in $\log p$ time:
\[
	\mathit{fexp}(b,p) = \left \{ \begin{array}{ll} 
		1 &  p = 0 \\ 
		\mathit{fexp}(b,p/2)^2 & p \text{ even} \\ 
		b \cdot \mathit{fexp}(b,p-1) & p \text{ odd} \end{array}
	\right .
\]

Without going into details of how recursion is handled, we implement staged fast exponentiation in \lang~below.  
We use {\tt \$} in place of $\fut$ in this an other examples.

\begin{lstlisting} 
let fexp (b : $int, e : int) : $int =
	if e == 0 then
		next{1}
	else if (e mod 2) == 0 then
		next{let x = prev{fexp(b,p/2)} in x*x}
	else
		next{prev{b} * prev{fexp(b,p-1)}}		
\end{lstlisting}

Although the code for {\tt fexp} looks very much like the stage-less mathematical definition above, it is in fact a staged program.  
See how the {\tt if}s and exponent decomposition are all at stage \bbone, since they occur within $\prev$ blocks.  
Essentially, the type system is sufficient to prove the observation that all of the decomposition of the exponent can be done at the stage \bbone~without requiring us to disrupt the elegant functional structure of the code.

This example also illustrates how stage \bbone~and stage \bbtwo~code can be arbitrarily nested.  Normal notions of eager evaluation reduce code in a leaf-to-root way, but this conflicts with the precept that we evaluate stage \bbone~before stage \bbtwo.  
To resolve this conflict, we need a formal semantics for \lang.
